---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pointdexter

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/cenuno/pointdexter.svg?branch=master)](https://travis-ci.org/cenuno/pointdexter)
<!-- badges: end -->

The package labels coordinate pairs located inside a polygon.

## Description

Two common spatial elements used in data analysis are points and polygons:

* points represent longitudinal and latitudinal coordinates, a unique pair that specifies a specific place; and

* polygons represent natural or artificial borders used to distinguish boundaries across geographies. 

The `pointdexter` package labels longitudinal and latitudinal coordinates located inside a polygon.

## Spatial Packages

The package is designed to work with both [sf](https://r-spatial.github.io/sf/) and [SpatialPolygonsDataFrame](https://www.rdocumentation.org/packages/sp/versions/1.2-5/topics/SpatialPolygonsDataFrame-class) objects.

By default, `pointdexter` only installs the [`sp`](https://www.rdocumentation.org/packages/sp/versions/1.3-1) package. However, it also works with the [`sf`](https://www.rdocumentation.org/packages/sf/versions/0.7-2) package.


## Installation

You can install the released version of `pointdexter` from [CRAN](https://CRAN.R-project.org) with:

```R
install.packages("pointdexter")
```

Development versions can be installed from GitHub:

```R
devtools::install_github("cenuno/pointdexter")
```

## Usage

### Build-in Data

`pointdexter` comes with built-in data - entirely due to the awesome and accessible [Chicago Data Portal](https://data.cityofchicago.org/) - to help you label points in polygons:

* Chicago Public Schools (CPS) - [School Profile Information, School Year (SY) 2018-2019](https://data.cityofchicago.org/Education/Chicago-Public-Schools-School-Profile-Information-/kh4r-387c)

* City of Chicago [boundary](https://data.cityofchicago.org/Facilities-Geographic-Boundaries/Boundaries-City/ewy2-6yfk)
    + as a SpatialPolygonsDataFrame; and
    + as a simple feature
    
* Chicago's [77 community areas](https://data.cityofchicago.org/Facilities-Geographic-Boundaries/Boundaries-Community-Areas-current-/cauq-8yn6)
    + as a SpatialPolygonsDataFrame; and
    + as a simple feature
    
### Example

Using built-in data, down below is an example of how to use `pointdexter` to label coordinate pairs with their appropriate polygon:

```{r example}
# load necessary packages ----
library(sp)          # classes and methods for spatial data
library(pointdexter) # label coordinate pairs in polygons
library(knitr)       # dynamic report generation

# load necessary data ----
data("cps_sy1819")
data("community_areas_spdf")

# create list of coordinate pair matrices for each community area ----
community.area.boundaries <-
  GetPolygonBoundaries(my.polygon = community_areas_spdf
                       , labels = community_areas_spdf$community)

# identify the community that each cps school lies in ----
cps_sy1819$community <-
  LabelPointsWithinPolygons(lng = cps_sy1819$school_longitude
                            , lat = cps_sy1819$school_latitude
                            , polygon.boundaries = community.area.boundaries)

# store relevant columns ----
relevant.columns <-
  c("school_id", "short_name"
    , "school_longitude", "school_latitude", "community")

# show first few records ----
kable(head(cps_sy1819[, relevant.columns], n = 10))
```

```{r visualize}
# plot only CPS high schools in the the Austin, Rogers Park, and West Elsdon community areas ----

# create filter condition ----
filter.condition <- 
  which(cps_sy1819$is_high_school &
          cps_sy1819$community %in% c("AUSTIN"
                                      , "ROGERS PARK"
                                      , "WEST ELSDON"))

# filter cps records to those that matched our condition ---
df <- cps_sy1819[filter.condition, ]

# note: clear plot space
par(mar = c(0, 0, 1, 0))
plot(x = community_areas_spdf
     , main = "CPS High Schools in Austin, Rogers Park, and West Elsdon communities, SY1819"
     , cex.main = 0.75
     , col = "gray85"
     , border = "dodgerblue4")
points(x = df$school_longitude
       , y = df$school_latitude
       , pch = 19
       , col = rgb(red = 212, green = 69, blue = 0
                   , alpha = 90
                   , maxColorValue = 255)
       , cex = 1)

# store relevant columns ----
relevant.columns <-
  c("school_id", "short_name", "is_high_school"
    , "school_longitude", "school_latitude", "community")

# print table of those schools that met the condition ----
kable(df[order(df$community), relevant.columns], row.names = FALSE)

```

## Resources

After you've installed the package, be sure to view the help files that introduce you to `pointdexter`'s two functions:

1. `?pointdexter::GetPolygonBoundaries()`; and
2. `?pointdexter::LabelPointsWithinPolygons()`

## Feedback

### Cite

If you use `pointdexter` for any analysis, I would love to hear about it! You can also cite the package according to `citation("pointdexter")`.

### Contribute

[Issues](https://github.com/cenuno/pointdexter/issues) and [pull requests](https://github.com/cenuno/pointdexter/pulls) are welcome anytime!
